---
title: "P8105_hw2"
author: "Nhu Nguyen"
date: "2023-09-27"
output: github_document
---
**load necessary libraries**
```{r}
library(tidyverse)
```
## problem 1

importing and cleaning dataset 

```{r}
pols_month = 
  read_csv("./fivethirtyeight_datasets/pols-month.csv") %>% 
  separate(mon, into = c("year", "month", "day")) %>% 
  mutate(month = recode(month, `01` = "Jan", `02` = "Feb", `03` = "Mar", `04` = "Apr", `05` = "May", `06` = "Jun", `07` = "Jul", `08` = "Aug", `09` = "Sep", `10` = "Oct", `11` = "Nov", `12` = "Dec")) %>% 
  mutate(president = case_when(
    prez_dem == 1 ~ "dem",
    prez_gop == 1 ~ "gop"
  )) %>% 
  select(-prez_dem, -prez_gop, -day) %>% 
  relocate (year, month, president) 

snp = 
  read_csv("./fivethirtyeight_datasets/snp.csv") %>% 
  separate(date, into = c("month", "day", "year")) %>% 
  mutate(month = recode(month, `1` = "Jan", `2` = "Feb", `3` = "Mar", `4` = "Apr", `5` = "May", `6` = "Jun", `7` = "Jul", `8` = "Aug", `9` = "Sep", `10` = "Oct", `11` = "Nov", `12` = "Dec")) %>% 
  mutate(year = ifelse(year >= 50, paste0("19", year), paste0("20", year))) %>% 
  relocate(year, month) %>% 
  select(-day) 

unemployment = 
  read_csv("./fivethirtyeight_datasets/unemployment.csv") %>% 
  pivot_longer(
    Jan:Dec,
    names_to = "month"
  ) %>% 
  rename_with(tolower, everything()) 
```
converting variable year from numeric to character data type in unemployment dataset
```{r}
unemployment$year <- as.character(unemployment$year)
```

merging snp_df and pols_month_df
```{r}
pols_snp_merge = left_join(pols_month, snp, by = c("year", "month"))
```

merging unemployment_df into our merged dataset
```{r}
fivethirtyeight_merge = full_join(pols_snp_merge, unemployment, by = c("year", "month"), copy = TRUE)
```

**description:**
The pol_month data set contains 822 observations of 9 variables regarding number republican or democratic national politicians associated with the month & year. The snp data set contains 787 observations of 3 variables relating to the closing value of the S&P stock index associated month & year. The unemployment data set contains 816 observations of 3 variables pertaining to the percentage of unemployment associated with the month & year. When merging all three data sets together, it creates a data set that contains 828 observations of 12 variables, with information pertaining to unemployment percentage and S&P stock index closing value, and whether the president during the associated month & date was republican or democrat. The relevant years with no missing information is from Jan 1950- Jun 2015. The key variables to look at would be the year, president, close, and value. 

## problem 2 
```{r}
trash =
  readxl::read_excel("./Trash_Wheel_Collection_Data.xlsx", range = "2A:)
```


