---
title: "P8105_hw2"
author: "Nhu Nguyen"
date: "2023-09-27"
output: github_document
---
load necessary libraries
```{r}
library(tidyverse)
```
## problem 1

importing and cleaning dataset 

```{r}
pols_month = 
  read_csv("./fivethirtyeight_datasets/pols-month.csv") %>% 
  separate(mon, into = c("year", "month", "day")) %>% 
  mutate(
    month = recode(month, `01` = "Jan", `02` = "Feb", `03` = "Mar", `04` = "Apr", `05` = "May", `06` = "Jun", `07` = "Jul", `08` = "Aug", `09` = "Sep", `10` = "Oct", `11` = "Nov", `12` = "Dec"), 
  president = case_when(
    prez_dem == 1 ~ "dem",
    prez_gop == 1 ~ "gop"
  ),
  year = as.integer(year)) %>% 
  select(-prez_dem, -prez_gop, -day) %>% 
  relocate (year, month, president) 

snp = 
  read_csv("./fivethirtyeight_datasets/snp.csv") %>% 
  separate(date, into = c("month", "day", "year")) %>% 
  mutate(
    month = recode(month, `1` = "Jan", `2` = "Feb", `3` = "Mar", `4` = "Apr", `5` = "May", `6` = "Jun", `7` = "Jul", `8` = "Aug", `9` = "Sep", `10` = "Oct", `11` = "Nov", `12` = "Dec"),
  year = ifelse(year >= 50, paste0("19", year), paste0("20", year)),
  year = as.integer(year)) %>% 
  relocate(year, month) %>% 
  select(-day) 

unemployment = 
  read_csv("./fivethirtyeight_datasets/unemployment.csv") %>% 
  pivot_longer(
    Jan:Dec,
    names_to = "month"
  ) %>% 
  rename_with(tolower, everything()) 
```

merging snp and pols_month
```{r}
pols_snp_merge = left_join(pols_month, snp, by = c("year", "month"))
```

merging unemployment into our merged dataset
```{r}
fivethirtyeight_merge = full_join(pols_snp_merge, unemployment, by = c("year", "month"), copy = TRUE)
```

**description**
The pols_month data set contains `r nrow(pols_month)` observations of `r ncol(pols_month)` variables regarding number republican or democratic national politicians associated with the month & year. The snp data set contains `r nrow(snp)` observations of `r ncol(snp)` variables relating to the closing value of the S&P stock index associated month & year. The unemployment data set contains `r nrow(unemployment)` observations of `r ncol(unemployment)` variables pertaining to the percentage of unemployment associated with the month & year. When merging all three data sets together, it creates a data set that contains `r nrow(fivethirtyeight_merge)` observations of `r ncol(fivethirtyeight_merge)` variables, with information pertaining to unemployment percentage and S&P stock index closing value, and whether the president during the associated month & date was republican or democrat. The relevant years with no missing information is from Jan 1950- Jun 2015. The key variables to look at would be the year, `president`, `close`, and `value`.

## problem 2 ##

importing and cleaning data, making sure that all variable names are labeled the same for easy merge

```{r}
trash =  
  readxl::read_excel("./TrashWheelCollectionData.xlsx", sheet = 1, range = "A3:N586", col_names = c("dumpster", "month", "year", "date", "weight_tons", "volume (cubic yards)", "plastic_bottles", "polystyrene", "cigarette_butts", "glass_bottles", "plastic_bags", "wrappers", "sports", "homes_powered")) %>% 
  janitor::clean_names() %>% 
  mutate(homes_powered = (
    (weight_tons*500)/1000), 
    trash_var = "mr.trash_wheel", 
    year = as.integer(year)) 

trash_wheel =
  readxl::read_excel("./TrashWheelCollectionData.xlsx", sheet = 2, range = "A3:M108", col_names = c("dumpster", "month", "year", "date", "weight_tons", "volume (cubic yards)", "plastic_bottles", "polystyrene", "cigarette_butts", "glass_bottles", "plastic_bags", "wrappers", "homes_powered")) %>% 
  janitor::clean_names() %>% 
  mutate(homes_powered = (
    (weight_tons*500)/1000), 
    year = as.integer(year), 
    trash_var = "prof_trash_wheel") 

gwynnda =
  readxl::read_excel("./TrashWheelCollectionData.xlsx", sheet = 4, range = "A3:L157", col_names = c("dumpster", "month", "year", "date", "weight_tons", "volume (cubic yards)", "plastic_bottles", "polystyrene", "cigarette_butts",  "plastic_bags", "wrappers", "homes_powered")) %>% 
  janitor::clean_names() %>% 
  mutate(homes_powered = (
    (weight_tons*500)/1000), 
    year = as.integer(year), 
    trash_var = "gwynnda")
```

merging all data set: 

```{r}
trash_tidy = 
  bind_rows(trash, trash_wheel, gwynnda) %>% 
  janitor::clean_names() %>% 
  relocate(trash_var)
```

cigarette butts collected by Gwynnda in July 2021
```{r}
gwynnda_july <- subset(gwynnda, month == "July" & year == "2021")
```

**description:**
This data set contains estimated trash gathered from a water-wheel vessel that removes trash from the Inner Harbor in Baltimore, Maryland. It contains `r nrow(trash_tidy)` observations and `r ncol(trash_tidy)` variables. Some key variables in this data set includes `weight_tons`, `volumne_cubic_yards`, and `homes_powered`. The total weight of trash collected by Professor Trash Wheel was `r sum(trash_wheel$weight_tons)` tons. The total number of cigarette butts collected by Gwynnda in July 2021 was `r sum(gwynnda_july$cigarette_butts)`

## problem 3

importing baseline data

```{r}
baseline = 
  read_csv("./data_mci/MCI_baseline.csv", skip = 1)%>%
  janitor::clean_names() %>% 
  mutate(
    sex = recode(sex, `1` = "male", `0` = "female"),
    apoe4 = recode(apoe4, `1` = "carrier", `0` = "non-carrier")) %>% 
    filter(age_at_onset > current_age | age_at_onset == ".")
```

proportion of women in the study who are apoe4 carriers: 
```{r}
baseline %>% 
  group_by(sex) %>% 
  summarize(prop_apoe4_carrier = mean(apoe4 == "carrier", na.rm = TRUE)) %>% 
  filter(sex == "female")
```


**description:**
Because the first row in the data set was not the variables, but was information of what the numeric values indicate, I had to specify to R to skip the first row and read the second row as the variable column. Additionally, I re-coded sex so that the output would either be female or male and apoe4 as carrier or non-carrier since they were numeric values in the raw dataset. I then excluded those who's `age_at_onset` predates `current_age` as they did not meet the inclusion criteria. The average baseline age in this study was `r mean(baseline$current_age)` years old. The proportion of women in the study who are APOE4 carriers is 30%.

importing longitudinally observed biomarker values:
```{r}
amyloid = 
  read_csv("./data_mci/mci_amyloid.csv", skip = 1) %>% 
  janitor::clean_names() %>% 
  pivot_longer(
    baseline:time_8,
    names_to = "time",
    names_prefix = "time_"
  ) %>% 
  mutate(
    id = study_id, 
    time = recode(time, "baseline" = "0"),
    id = as.integer(id)) %>% 
  select (id, everything(), -study_id)
```
**description:**
Similar to the baseline data set, the amyloid data set's first row gave information of the units, thus I had to specify to R to skip the first row and read the second row as the variable. I then created a new variable `time` and moved the columns into that variable through using the pivot_longer function so that I will be able to merge the dataset. I also had to change the variable `study_id` to `id` so that it would match the baseline data variables. 

checking to see if some participants only appear in baseline or amyloid data set 
```{r}
mci_2 <- anti_join(baseline, amyloid, by = "id" )
```
There are `r nrow(mci_2)` participants that only appear in only the baseline or amyloid dataset. 

merging data set together:
```{r}
mci_merge = inner_join(baseline, amyloid, by = "id")
```

**description:**
In the mci_merge data set, I did an inner_join of the baseline and amyloid data set by variable `id` to ensure that only that appear in both datasets are retained. After merging, there are `r nrow(mci_merge)/5` participants left in the study. 

exporting results as csv onto desktop 
```{r}
p8105_hw2_nn2477 <- "/Users/nhunguyen/Desktop/P8150/p8105_hw2_nn2477/mci_merge.csv"
write.csv(mci_merge, file = p8105_hw2_nn2477, row.names = FALSE)
```

